---
title: "MitoSignatures"
author: "Isabel Duarte"
date: "2023-05-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warnings=FALSE, cache = TRUE)

library(here)
library(tidyverse)
library(edgeR)

```

## Differential expression | GTEx

```{r gtex, eval=FALSE}

# Create list to hold GTEx objects
gtex_de <- list()

# Load data and sample metadata
gtex_de$data_clean <- readRDS(here("data/processed/gtex_data_clean.RDS"))

# Ensure the sample metadata matches the identity and order of the columns in the expression data

if (identical(colnames(gtex_de$data_clean$combined_counts)[-1],
              gtex_de$data_clean$metadata$sample_id)) {
  print("PASS | The identity and order of the vectors are the same.")
} else {
  print("REJECT | The identity and order of the vectors are different.")
}

# Check how many rows have at least one NA (NAs are not permitted by DGEList function)
gtex_de$counts_NAs <- gtex_de$data_clean$combined_counts[complete.cases(gtex_de$data_clean$combined_counts) == FALSE, ]
  
# Transform count data-frame to matrix with row names required by EdgeR
# and remove NAs (if they exist)
gtex_de$counts_matrix <- gtex_de$data_clean$combined_counts %>%
  na.omit() %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()

#
##
### edgeR analysis
##
#

# Create a DGEList object
gtex_de$dge_data <- DGEList(counts = gtex_de$counts_matrix)


# Set up the design matrix | Estimate coefficients for each organ independently

# gtex_de$design <- model.matrix(~ 0 + organ,
#                        data = gtex_de$data_clean$metadata)

gtex_de$design <- model.matrix(~ 0 + metabolic_rate,
                       data = gtex_de$data_clean$metadata)

# Set up the contrasts matrix

# # Make the contrasts for high versus low metabolic-rate organs:
# # requires the sum of the coefficients for each metabolic-rate level
# # divided by 3 (average) | high - low
# gtex_de$contrasts <- limma::makeContrasts(
#   (organbrain+organkidney+organliver)/3 -
#     (organbladder+organcolon+organskin)/3,
#   levels=colnames(gtex_de$design)
# )
# # Change column names
# colnames(gtex_de$contrasts) <- c("brain_kidney_liver_VS_bladder_colon_skin")

# Make the contrasts for low versus high metabolic rate
gtex_de$contrasts <- limma::makeContrasts(
  metabolic_ratehigh - metabolic_ratelow,
  levels=colnames(gtex_de$design)
)
# Change column names
colnames(gtex_de$contrasts) <- c("metab_high_VS_metab_low")


# Filter low-expression genes | Requires a gene to have a normalized count of at least 10 in at least k samples, where k is the smallest group size
gtex_de$keep <- filterByExpr(gtex_de$dge_data,
                                   design = gtex_de$design)
gtex_de$dge_data_filtered <- gtex_de$dge_data[gtex_de$keep, , keep.lib.sizes=FALSE]

# Perform Library Size Normalization | Slow step
gtex_de$dge_data_filtered <- calcNormFactors(gtex_de$dge_data_filtered) # more recent edgeR uses normLibSizes()

# Estimate dispersions | Slow step
gtex_de$dge_data_filtered <- estimateDisp(gtex_de$dge_data_filtered,
                                                design = gtex_de$design)

# Perform likelihood ratio tests
# Fit the negative binomial generalized log-linear model
gtex_de$fit <- glmFit(gtex_de$dge_data_filtered,
                            design=gtex_de$design,
                   contrast = gtex_de$contrasts)

# Perform likelihood ratio tests | Slow step
gtex_de$lrt <- glmLRT(gtex_de$fit)   

# Get the topTags
gtex_de$topGenes <- topTags(gtex_de$lrt, n=NULL,
                         adjust.method = "BH", 
                         sort.by = "PValue", 
                         p.value = 0.05)

# Save the diff expression analysis results
saveRDS(gtex_de, file = here("data/analysis/gtex_de.RDS"))

```


## Differential expression | TCGA


```{r tcga, eval=FALSE}

# Create list to hold TCGA objects
tcga_de <- list()

# Load data and sample metadata
tcga_de$data_clean <- readRDS(here("data/processed/tcga_data_clean.RDS"))

# Ensure the sample metadata matches the identity and order of the columns in the expression data

if (identical(colnames(tcga_de$data_clean$combined_counts)[-1],
              tcga_de$data_clean$metadata$sample_id)) {
  print("PASS | The identity and order of the vectors are the same.")
} else {
  print("REJECT | The identity and order of the vectors are different.")
}

# Check how many rows have at least one NA (NAs are not permitted by DGEList function)
tcga_de$counts_NAs <- tcga_de$data_clean$combined_counts[complete.cases(tcga_de$data_clean$combined_counts) == FALSE, ]
  
# Transform count data-frame to matrix with row names required by EdgeR
# and remove NAs (if they exist)
tcga_de$counts_matrix <- tcga_de$data_clean$combined_counts %>%
  na.omit() %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()
  
#
##
### edgeR analysis
##
#

# Create a DGEList object
tcga_de$dge_data <- DGEList(counts = tcga_de$counts_matrix)

# Set up the design matrix

tcga_de$design <- model.matrix(~ 0 + metabolic_rate,
                            data = tcga_de$data_clean$metadata)

# tcga_de$design <- model.matrix(~ 0 + organ,
#                             data = tcga_de$data_clean$metadata)

# Set up the contrasts matrix

# Make the contrasts for high versus low metabolic-rate organs:
tcga_de$contrasts <- limma::makeContrasts(
  metabolic_ratehigh - metabolic_ratelow,
  levels=colnames(tcga_de$design)
)
# Change column names
colnames(tcga_de$contrasts) <- c("metab_high_VS_metab_low")


# # Make the contrasts for high versus low metabolic-rate organs:
# # requires the sum of the coefficients for each metabolic-rate level
# # divided by 3 (average) | high - low
# tcga_de$contrasts <- limma::makeContrasts(
#   (organbrain+organkidney+organliver)/3 -
#     (organbladder+organcolon+organskin)/3,
#   levels=colnames(tcga_de$design)
# )
# # Change column names
# colnames(tcga_de$contrasts) <- c("brain_kidney_liver_VS_bladder_colon_skin")


# Filter low-expression genes
tcga_de$keep <- filterByExpr(tcga_de$dge_data,
                          design = tcga_de$design)
tcga_de$dge_data_filtered <- tcga_de$dge_data[tcga_de$keep, , keep.lib.sizes=FALSE]

# Perform Library Size Normalization | Slow step
tcga_de$dge_data_filtered <- calcNormFactors(tcga_de$dge_data_filtered)

# Estimate dispersions | Slow step
tcga_de$dge_data_filtered <- estimateDisp(tcga_de$dge_data_filtered,
                                       design = tcga_de$design)

### To perform likelihood ratio tests
# Fit the negative binomial generalized log-linear model
tcga_de$fit <- glmFit(tcga_de$dge_data_filtered,
                   design=tcga_de$design,
                   contrast = tcga_de$contrasts)

# Perform likelihood ratio tests
tcga_de$lrt <- glmLRT(tcga_de$fit)

# Extract the differentially expressed genes
tcga_de$topGenes <- topTags(tcga_de$lrt, n=NULL,
                         adjust.method = "BH", 
                         sort.by = "PValue", 
                         p.value = 0.05)

# Save the diff expression analysis results
saveRDS(tcga_de, file = here("data/analysis/tcga_de.RDS"))

```


## Differential expression | Combined TCGA and GTEx | Model ~ 0 + metabolic_rate + cancer_status 

```{r combined, eval=FALSE}

# Create list to hold analysis objects
combined_de_no_interaction <- list()

# Load data and sample metadata
combined_de_no_interaction$data_clean$combined_counts <- readRDS(here("data/processed/combined_counts_tcga_gtex.RDS"))
combined_de_no_interaction$data_clean$metadata <- readRDS(here("data/processed/combined_metadata_tcga_gtex.RDS"))

# Ensure the sample metadata matches the identity and order of the columns in the expression data

if (identical(colnames(combined_de_no_interaction$data_clean$combined_counts)[-1],
              combined_de_no_interaction$data_clean$metadata$sample_id)) {
  print("PASS | The identity and order of the vectors are the same.")
} else {
  print("REJECT | The identity and order of the vectors are different.")
}

# Check how many rows have at least one NA (NAs are not permitted by DGEList function)
combined_de_no_interaction$counts_NAs <- combined_de_no_interaction$data_clean$combined_counts[complete.cases(combined_de_no_interaction$data_clean$combined_counts) == FALSE, ]

# Transform count data-frame to matrix with row names required by EdgeR
# and remove NAs (if they exist)
combined_de_no_interaction$counts_matrix <- combined_de_no_interaction$data_clean$combined_counts %>%
  na.omit() %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()

#
##
### edgeR analysis
##
#

# Create a DGEList object
combined_de_no_interaction$dge_data <- DGEList(counts = combined_de_no_interaction$counts_matrix)


#
##
### Model design without interaction terms
##
#

# Set up the design matrix
combined_de_no_interaction$design <- model.matrix(~ 0 + metabolic_rate + cancer_status,
                                                  data = combined_de_no_interaction$data_clean$metadata)

# Set up the contrasts matrix
combined_de_no_interaction$contrasts <- limma::makeContrasts(
  (metabolic_ratehigh - metabolic_ratelow),
  levels=colnames(combined_de_no_interaction$design)
)

# # Change column names
colnames(combined_de_no_interaction$contrasts) <- c("MetRateHigh_vs_MetRateLow")


# Filter low-expression genes
combined_de_no_interaction$keep <- filterByExpr(combined_de_no_interaction$dge_data,
                                                design = combined_de_no_interaction$design)
combined_de_no_interaction$dge_data_filtered <- combined_de_no_interaction$dge_data[combined_de_no_interaction$keep, , keep.lib.sizes=FALSE]

# Perform Library Size Normalization | Slow step
combined_de_no_interaction$dge_data_filtered <- calcNormFactors(combined_de_no_interaction$dge_data_filtered)

# Estimate dispersions | Slow step
combined_de_no_interaction$dge_data_filtered <- estimateDisp(combined_de_no_interaction$dge_data_filtered,
                                                             design = combined_de_no_interaction$design)

### To perform likelihood ratio tests
# Fit the negative binomial generalized log-linear model
combined_de_no_interaction$fit <- glmFit(combined_de_no_interaction$dge_data_filtered,
                                         design=combined_de_no_interaction$design,
                                         contrast = combined_de_no_interaction$contrasts)

# Perform likelihood ratio tests
combined_de_no_interaction$lrt <- glmLRT(combined_de_no_interaction$fit)

# Extract the differentially expressed genes
combined_de_no_interaction$topGenes <- topTags(combined_de_no_interaction$lrt, n=NULL,
                                               adjust.method = "BH", 
                                               sort.by = "PValue", 
                                               p.value = 0.05)


### Save the diff expression analysis results
saveRDS(combined_de_no_interaction, file = here("data/analysis/combined_de_no_interaction.RDS"))

```



# NOTE - CODE AFTER THIS SECTION IS STILL UNDER DEVELOPMENT

## Differential expression | Combined TCGA and GTEx | Model ~ 0 + metabolic_rate:cancer_status


```{r combined1, eval=FALSE}

# Create list to hold analysis objects
combined_de_only_interaction <- list()

# Load data and sample metadata
combined_de_only_interaction$data_clean$combined_counts <-
  readRDS(here("data/processed/combined_counts_tcga_gtex.RDS"))
combined_de_only_interaction$data_clean$metadata <-
  readRDS(here("data/processed/combined_metadata_tcga_gtex.RDS"))

# Ensure the sample metadata matches the identity and order of the columns in the expression data
if (identical(
  colnames(combined_de_only_interaction$data_clean$combined_counts)[-1],
  combined_de_only_interaction$data_clean$metadata$sample_id
)) {
  print("PASS | The identity and order of the vectors are the same.")
} else {
  print("REJECT | The identity and order of the vectors are different.")
}

# Transform count data-frame to matrix with row names required by EdgeR
# and remove NAs (if they exist)
combined_de_only_interaction$counts_matrix <-
  combined_de_only_interaction$data_clean$combined_counts %>%
  na.omit() %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()

#
##
### edgeR analysis
##
#

# Create a DGEList object
combined_de_only_interaction$dge_data <-
  DGEList(counts = combined_de_only_interaction$counts_matrix)


#
##
### Model design with interaction term between metabolic_rate and cancer_status
##
#

# Set up the design matrix
combined_de_only_interaction$design <-
  model.matrix(
    ~ 0 + metabolic_rate:cancer_status,
    data = combined_de_only_interaction$data_clean$metadata
  )

# Correct the names of the design column (R does not allow : in the names)
colnames(combined_de_only_interaction$design) <- 
  sub(":", "_",
      colnames(combined_de_only_interaction$design)) %>%
  sub("metabolic_rate", "", .) %>%
  sub("cancer_status", "", .)
  

# Set up the contrasts matrix
combined_de_only_interaction$contrasts <- limma::makeContrasts(
  (high_non_cancer - low_non_cancer) - (high_cancer - low_cancer),
  levels = colnames(combined_de_only_interaction$design)
)

# # Change column names
colnames(combined_de_only_interaction$contrasts) <-
  c("highNonCancer_lowNonCancer_VS_highCancer_lowCancer")


# Filter low-expression genes
combined_de_only_interaction$keep <-
  filterByExpr(combined_de_only_interaction$dge_data,
               design = combined_de_only_interaction$design)
combined_de_only_interaction$dge_data_filtered <-
  combined_de_only_interaction$dge_data[combined_de_only_interaction$keep, , keep.lib.sizes =
                                          FALSE]

# Perform Library Size Normalization | Slow step
combined_de_only_interaction$dge_data_filtered <-
  calcNormFactors(combined_de_only_interaction$dge_data_filtered)

# Estimate dispersions | Slow step
combined_de_only_interaction$dge_data_filtered <-
  estimateDisp(combined_de_only_interaction$dge_data_filtered,
               design = combined_de_only_interaction$design)


### Perform likelihood ratio tests

# Fit the negative binomial generalized log-linear model USING CONTRAST MATRIX
combined_de_only_interaction$fit <- glmFit(combined_de_only_interaction$dge_data_filtered,
                                           design=combined_de_only_interaction$design,
                                           contrast = combined_de_only_interaction$contrasts) # ATTENTION: contrast matrix USED

# Perform likelihood ratio tests
combined_de_only_interaction$lrt <-
  glmLRT(combined_de_only_interaction$fit)

# Extract the differentially expressed genes
combined_de_only_interaction$topGenes <-
  topTags(
    combined_de_only_interaction$lrt,
    n = NULL,
    adjust.method = "BH",
    sort.by = "PValue",
    p.value = 0.05
  )


# Save the diff expression analysis results
saveRDS(
  combined_de_only_interaction,
  file = here("data/analysis/combined_de_only_interaction.RDS")
)

```


## Differential expression | Combined TCGA and GTEx | Model ~ 0 + metabolic_rate + cancer_status + metabolic_rate:cancer_status


```{r combined2, eval=TRUE}

# Create list to hold analysis objects
combined_de_with_interaction <- list()

# Load data and sample metadata
combined_de_with_interaction$data_clean$combined_counts <-
  readRDS(here("data/processed/combined_counts_tcga_gtex.RDS"))
combined_de_with_interaction$data_clean$metadata <-
  readRDS(here("data/processed/combined_metadata_tcga_gtex.RDS"))

# Ensure the sample metadata matches the identity and order of the columns in the expression data

if (identical(
  colnames(combined_de_with_interaction$data_clean$combined_counts)[-1],
  combined_de_with_interaction$data_clean$metadata$sample_id
)) {
  print("PASS | The identity and order of the vectors are the same.")
} else {
  print("REJECT | The identity and order of the vectors are different.")
}

# Transform count data-frame to matrix with row names required by EdgeR
# and remove NAs (if they exist)
combined_de_with_interaction$counts_matrix <-
  combined_de_with_interaction$data_clean$combined_counts %>%
  na.omit() %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()

#
##
### edgeR analysis
##
#

# Create a DGEList object
combined_de_with_interaction$dge_data <-
  DGEList(counts = combined_de_with_interaction$counts_matrix)


#
##
### Model design with interaction term between metabolic_rate and cancer_status
##
#

# combined_de_with_interaction$design <-
#   model.matrix(
#     ~ 0 + metabolic_rate +
#       cancer_status +
#       cancer_status:metabolic_rate,
#     data = combined_de_with_interaction$data_clean$metadata
#   )
# 
# # Correct the names of the design column (R does not allow : in the names)
# colnames(combined_de_with_interaction$design) <- sub(":", "_",
#                                                      colnames(combined_de_with_interaction$design))
# 
# # Set up the contrasts matrix
# combined_de_with_interaction$contrasts <- limma::makeContrasts(
#   (metabolic_ratehigh - metabolic_ratelow),
#   levels = colnames(combined_de_with_interaction$design)
# )
# 
# # # Change column names
# colnames(combined_de_with_interaction$contrasts) <-
#   c("MetRateHigh_vs_MetRateLow")


combined_de_with_interaction$design <-
  model.matrix(
    ~ 0 + metabolic_rate:cancer_status,
    data = combined_de_with_interaction$data_clean$metadata
  )

# Correct the names of the design column (R does not allow : in the names)
colnames(combined_de_with_interaction$design) <- 
  sub(":", "_",
      colnames(combined_de_with_interaction$design)) %>%
  sub("metabolic_rate", "", .) %>%
  sub("cancer_status", "", .)
  

# Set up the contrasts matrix
combined_de_with_interaction$contrasts <- limma::makeContrasts(
  (high_non_cancer - low_non_cancer) - (high_cancer - low_cancer),
  levels = colnames(combined_de_with_interaction$design)
)

# # Change column names
colnames(combined_de_with_interaction$contrasts) <-
  c("highNonCancer_lowNonCancer_VS_highCancer_lowCancer")


# Filter low-expression genes
combined_de_with_interaction$keep <-
  filterByExpr(combined_de_with_interaction$dge_data,
               design = combined_de_with_interaction$design)
combined_de_with_interaction$dge_data_filtered <-
  combined_de_with_interaction$dge_data[combined_de_with_interaction$keep, , keep.lib.sizes =
                                          FALSE]

# Perform Library Size Normalization | Slow step
combined_de_with_interaction$dge_data_filtered <-
  calcNormFactors(combined_de_with_interaction$dge_data_filtered)

# Estimate dispersions | Slow step
combined_de_with_interaction$dge_data_filtered <-
  estimateDisp(combined_de_with_interaction$dge_data_filtered,
               design = combined_de_with_interaction$design)

### To perform likelihood ratio tests

# Fit the negative binomial generalized log-linear model USING COEFFICIENT CONTRAST
# combined_de_with_interaction$fit <-
#   glmFit(
#     combined_de_with_interaction$dge_data_filtered,
#     design = combined_de_with_interaction$design,
#     contrast = metabolic_ratehigh_cancer_statuscancer
#   ) # ATTENTION: contrast matrix NOT used


# Fit the negative binomial generalized log-linear model USING CONTRAST MATRIX
combined_de_with_interaction$fit <- glmFit(combined_de_with_interaction$dge_data_filtered,
                                           design=combined_de_with_interaction$design,
                                           contrast = combined_de_with_interaction$contrasts) # ATTENTION: contrast matrix USED


# Perform likelihood ratio tests
combined_de_with_interaction$lrt <-
  glmLRT(combined_de_with_interaction$fit)

# Extract the differentially expressed genes
combined_de_with_interaction$topGenes <-
  topTags(
    combined_de_with_interaction$lrt,
    n = NULL,
    adjust.method = "BH",
    sort.by = "PValue",
    p.value = 0.05
  )


# Save the diff expression analysis results
saveRDS(
  combined_de_with_interaction,
  file = here("data/analysis/combined_de_with_interaction.RDS")
)

```

